Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-11 rebuttal.tex

\comment{The authors state that 3D information is lost when only
  considering the molecular graph. However, they do not make it clear
  that the 3D space is implicitly encoded in the molecular graph –
  this is essential to cover properly.}
\response{We have update the text on page 2 to note that 3D
  information is implicit in 2D representations}

\comment{If this manuscript were to be published it would be
  completely rewriting the history of cheminformatics}
\response{While we agree with the reviewer that some key milestones in
  the history of cheminformatics were not included in the original
  article, we believe that the addition of the supplemental history
  addresses this problem. Furthermore, our original aim was \emph{not}
  to provide a complete historical overview of the field. Others have
  already provided such overviews. Our aim was to present a brief
  overview of \emph{topics} in cheminformatics, selected because we
  believed that they represent on going challenges for the field and
  thus amenable to coss-disciplinary efforts.}

\fbox{\textbf{Reviewer 3}}

The article over-simplifies concepts, such as molecular representations, for an audience which is perhaps even more well versed in graph theory than many cheminformaticians. The casual reader should be left with an understanding of specific challenges to which they can bring to bear methods from their field. Vague challenges are mentioned, but not in any clarity enough to go away and tackle them.

The manuscript provides a very specific overview on the aspects of the open source community in cheminformatics. Therefore, I would recommend that this be clarified in the title to avoid confusion or the manuscript broadened to
cover the actual extent of the field, which is far vaster than is covered here.

The authors make no (or very little) mention of the founding principles of cheminformatics. Very few references cover the algorithms that have been developed and published.

The manuscript is poor in relevant citations with none of the classical references provided, such as Morgan 1965, Weininger 1988, Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.

Molecular docking is also not mentioned, but is regarded as one of the key methods applied in cheminformatics. Docking should be included, along with other molecular modelling methods such as pharmacophores, shape searching,
molecular dynamics, as a distinct section.

Proteochemometric methods are mentioned a couple of times but these are largely un-noticed by the community currently. I do not think that it is necessarily appropriate to include these as tried-and-tested methods in an over-arching review such as this.

\comment{RDKit is not mentioned at all in this article and is widely used as a
cheminformatics toolkit in industry and academia and free and
open-source (although from commercial beginnings).}
\response{We agree that this toolkit should have been included. We
  have updated the text to take note of RDKit, though we still
  primarily focus on the CDK and OpenBabel. However, Table 2
  includes RDKit in the feature comparison}

\comment{p1, c2, l31: it is not clear to this reviewer what software
  and algorithms have only recently been made available. Certainly,
  the referenced book to which they refer contains a large number of
  algorithms that have already been published in the primary
  literature and frequently many decades ago. If the open source
  community has provided true novelty then please articulate this,
  otherwise cite the primary references that report these
  discoveries.}
\response{The text has been updated to note that most of the
  algorithms in cheminformatics have been known since the mid 20th
  century, but that freely accessible toolkits implementing them are
  relatively recent}

### 2011-11-11 rebuttal.tex

The manuscript provides a very specific overview on the aspects of the open source community in cheminformatics. Therefore, I would recommend that this be clarified in the title to avoid confusion or the manuscript broadened to
cover the actual extent of the field, which is far vaster than is covered here.

The authors make no (or very little) mention of the founding principles of cheminformatics. Very few references cover the algorithms that have been developed and published.

The manuscript is poor in relevant citations with none of the classical references provided, such as Morgan 1965, Weininger 1988, Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.

\comment{Molecular docking is also not mentioned, but is regarded as
  one of the key methods applied in cheminformatics. Docking should be
  included, along with other molecular modelling methods such as
  pharmacophores, shape searching, molecular dynamics, as a distinct
  section.}  
\response{These methods are mentioned on page 5 (column
  1). However, we have chosen not to place a more in depth discussion
  of these methods in their own section in the main document for a few
  reasons. First, due to space limitations this would require us to
  compress or drop other sections which we believe to be more
  relevant. Second, we are of the view that dokcing, for example, is
  not really a cheminformatics technique, though it makes extensive
  use of cheminformatic concepts. Instead we place it under molecular
  modeling. We realize that this is a somewhat subjective
  ``classification'' of methods. However we do make explicit note of
  these methods in the supplemental document where we make a
  distinction between cheminformatics and related subject areas such
  as quantum mechanics and molecular modeling.}

\comment{Proteochemometric methods are mentioned a couple of times but
  these are largely un-noticed by the community currently. I do not
  think that it is necessarily appropriate to include these as
  tried-and-tested methods in an over-arching review such as this.}
\response{While it is true that proteochemometric methodologies are
  not as well known as say docking or pharmacophore modeling, we
  choose to include them since they are an example of a (statistical)
  QSAR technique that does explicitly include ligand and receptor features}

\comment{RDKit is not mentioned at all in this article and is widely
  used as a cheminformatics toolkit in industry and academia and free
  and open-source (although from commercial beginnings).}
\response{We agree that this toolkit should have been included. We
  have updated the text on page 7 to take note of RDKit, though we
  still primarily focus on the CDK and OpenBabel. However, Table 2
  includes RDKit in the feature comparison}

### 2011-11-11 paper.tex
Added more responses

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR' \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking, pharmacophore modeling and proteochemometric methods
have been designed to take into account both ligand and receptor
structures.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that such aspects be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Computer science techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem. More recently, simultaneous modeling of
receptor and ligand has been addressed by the proteochemometric
approach as first reported by Lapinsh \textit{et
  al.}~\cite{lapinsh2001}.

>>>>

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR' \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking, pharmacophore modeling and proteochemometric methods
have been designed to take into account both ligand and receptor
structures. The last method is an example of an extension of
statistical QSAR methods to simultaneous model the f receptor and
ligand as first reported by Lapinsh \textit{et  al.}~\cite{lapinsh2001}.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that such aspects be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Computer science techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.
